Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics

### 2023-02-23 03_marlin_paper.qmd
off to abalone

Abstract: Designing effective spatial management strategies is challenging because marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond endogenously to ecosystem changes in ways that depend upon ecological and policy context. We present a tool – marlin -- to efficiently simulate the bio-economic dynamics of marine systems, designed to support both management and research. We demonstrate `marlin’s` capabilities by focusing on two case studies on the conservation and food production impacts of marine protected areas (MPAs): a coastal coral reef and a pelagic tuna fishery. In the coastal coral reef example, we explore how heterogeneity in species distributions and fleet preferences can affect distributional outcomes of MPAs. In the pelagic case study, we demonstrate how our model can be used to assess the climate resilience of different MPA design strategies, as well as the climate sensitivity of different fishing fleets. This paper demonstrates how intermediate complexity simulation of coupled bio-economic dynamics can help communities predict and potentially manage trade-offs between conservation, fisheries yields, and distributional outcomes resulting from spatial management.

@@

We define a few commonly used terms here. *Yields* refers to the volume of catch from fishing activity. *Spawning stock biomass* refers to the total biomass of reproductively mature fish in the water, a function of numbers, weight, and sexual maturity at age. *Depletion* is a common term in the fisheries literature, and reflects here the *spawning stock biomass* in a given time step relative to *unfished spawning stock biomass*. So, a depletion value of 1 means that the population is unfished, a depletion value of 0 means the population is extinct.

@@

Revenues for each fleet in a time step are a function of the total
amount of each species caught and its price ($\Pi$). The amount
caught is a function of the selectivity at age of each species for each
fleet ($\alpha$), the fishing efficiency of the fleet for that that
species (also called "catchability", *q*), the amount of fishing effort
of the fleet in question in that patch in a given time step (*E*), and
the total fishing mortality at age (including all other fleets) for that
particular species in that patch and time (*u*).

@@

Total catches or yield *Y* are calculated through the Baranov equation
[@baranov1918], which appropriately accounts for the total instantaneous
mortality (both fishing and natural) and divides the total amount of the
population biomass (*b*) killed between natural (*m*) and fishing (*u*)
sources, with the amount of biomass killed through fishing assigned to
yield *Y*.

@@

This means that each species experiences a total mortality at age in a
time step in a patch, individual fractions of which are portioned off as
catches and subsequently revenues for each fleet. Given revenues *R*, we
then calculate profits $\phi$ based on revenues and costs. Costs *C* are
calculated as a function of $\gamma_f$ that the scales the total cost
per unit effort for fleet *f*. $\beta$ allows for the cost per unit
effort effort to scale non-linearly, and $\eta$ is an additional cost
per unit effort of fishing in each patch *p*, allowing for the model to
account for travel costs for different patches. Travel costs are
calculated based on the Euclidean distance of each patch to the nearest
port of a given fleet, and users can specify any number from zero to the
number of patches of port locations. When no ports are specified, travel
costs are zero.

@@

#### Tuning Fishing Fleets

The degree of fishing pressure exerted by a given set of fishing fleets
on each species is a function of a range of parameters including the
total amount of effort, the selectivity ogives, cost functions, the
spatial distribution of the resource, the relative prices across
species, and the catchability coefficient of each fleet for each
species. `marlin` provides a tuning option to help users achieve desired
biological outcomes from their fleets.

Users can tune their fleets in one of two ways. First, they can specify
a target exploitation rate *u* for each species in their simulation.
Taking all the other parameters of the model as given, `marlin` then
adjusts the catchability coefficients $q_{f,s}$ for each fleet *f* and
species *s* such that the equilibrium exploitation rate for each species
matches the desired level. Second, they can specify a target depletion,
defined as total spawning stock biomass under fishing divided by total
unfished spawning stock biomass, and the model will adjust the
catchability coefficients $q_{f,s}$ for each fleet *f* and species *s*
such that the equilibrium depletion for each species matches the desired
level.

@@

The underlying population model used is an age structured single-species
model in the manner of @ovando2021a. The population model requires many
parameters. However, if the user supplies either a scientific or common
name for the species in question (scientific preferred), the model will
supply default values for that species based on the values reported in
FishLife [internet connection required, @thorson2020]. All of the life history values can
alternatively be replaced with user-supplied values as needed and should be for more tactical applications; the default values simply ensure that reasonable and internally consistent
values for the species in question are supplied as a starting point.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of *advection* (follow currents), and *taxis* (active movement towards preferred habitat) and *diffusion* (essentially remaining variation in movement not explained by advection or taxis). For now, `marlin` focuses just on the
diffusion and taxis components of this model, assuming that advection is zero, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way, `marlin` allows users to run anything from a simple Gaussian
dispersal kernel up to a system governed by species that passively
diffuse around a dynamic thermal range.

We provide a brief overview of the the general CTMC method here, see
@thorson2021a for a detailed description. Under this framework, movement
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of the sum of the diffusion $\boldsymbol{D}$ and taxis $\boldsymbol{\tau}$ matrices scaled by the width of the time step $\Delta_{t}$ and the length of the edge of each cell $\Delta_d$ that the model is running under. So for example, if we are running a model on a monthly timestep, and we define the units of time in years, then $\Delta_{t} = 1/12$. Similarly, if we are running a model where one cell equals 100km^2^, then $\Delta_d = 100km$. This parameterization allows users to set the effective area of the spatial domain through two avenues; the number of cells, which effectively scales the resolution of the model, and the area of each cell, which scales the spatial extent of the simulation.

@@

The individual components of the diffusion matrix are filled based on an adjacency matrix, which defines whether two cells are both adjacent and water (as opposed to land or another physical barrier), and a diffusion rate $D$ defined in units of length of the side of a cell per unit of time. For example, if we are defining the time units as years and the distance units as kilometers, for a tuna $D$ might be 1,000 $\frac{KM}{Year}$. The $\Delta_{t}$ and $\Delta_{p}$ parameters then translate the movement rate $D$ to match the time step and cell size used in a simulation. This "scale free" parameterization means that appropriate value of $D$ can be identified for a species and then set, regardless of the time step or cell size used in the simulation model itself.

@@

The taxis matrix $\boldsymbol{\tau}$ is built off of the difference in log habitat quality between each patch, the adjacency matrix, and temporal and spatial resolution of the model ($\Delta_{t}$ and $\Delta_{p}$), as well as a scalar term $\lambda$. $\lambda$ scales the magnitude of the habitat
gradient between patches relative to the diffusion
parameter $D$. When the diffusion and taxis matrices are estimated
jointly from actual data in the manner of @thorson2021a, $\lambda$
should be set to zero. However, when users are manually creating both
the diffusion and taxis matrices for simulation, the relative strength of diffusion to
taxis must be defined by the user. Setting higher values of $\lambda$
allows users to simulate a scenario where taxis is much more significant
than diffusion.

$$
\tau_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}}(1 + \lambda d_{s,f})e^{h(p2,t,s,a) - h(p1,t,s,a)} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} \tau_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

For both the diffusion and taxis matrix, we allow for the inclusion of
physical barriers to movement (i.e. land). Pairs of patches that are
adjacent but in which one or both patches are a barrier to movement are
set as non-adjacent. The CTMC model then produces movement dynamics that
move around barriers rather than over them. This is different than
setting habitat in a patch to zero; an animal will not preferentially move
towards a patch with zero habitat all else being equal, but can move
over a zero habitat patch towards another better patch. They most move around a physical barrier.

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel. However, the time component *t* in @eq-movement allows
the movement model to be run on arbitrarily fine time steps. Second, the
taxis model allow for more active habitat choices by species, allowing
us to simulate both preferences of species in space and time
efficiently. Lastly, this form has the advantage that the parameters are
directly estimable from real data. So, if provided with say spatial
abundance data and a tagging study, users can estimate the movement
parameters in the same manner as @thorson2021a, and then pass their
estimated parameters to `marlin` for simulation.

@@

where *BH* is the Beverton-Holt recruitment function, *ssb* is
spawning-stock-biomass, $l_a$ is the mean length at at age, $l_{sel}$ is
the length at which on average 50% of individuals are selected by the
fishery, and $\delta_{sel}$ are the additional units of length at which
on average 95% of fish are selected by the fishery.

Spawning stock biomass *ssb* is calculated by converting age to mean
length, calculating weight at age, maturity at age, and then calculating
spawning stock biomass as the sum of spawning potential at age in a
given time step, taking into account the potential for hyperallometry.
Age is converted to length through the von Bertalanffy growth equation,
assuming log-normally distributed variation in the length at age.

@@

Where *u* is a log-normally distributed term allowing for a constant
coefficient of variation in the length at age relationship

@@

Biomass *b* at age is then given by the weight at length equation
governed by a scaling coefficient $wa_s$ and an exponent $wb_s$ that
controls the rate at which volume scales with length

$$
  b_{a,s} = wa_s \times l_{a,s}^{wb_s}
$$ {#eq-weight}

Maturity is then calculated as a logistic function where $l_{mat}$ is
the length at which on average 50% of individuals are sexually mature,
and $\delta_{mat}$ is the unit of length beyond $l_{mat}$ at which on
average 95% of fish are sexually mature.

@@

Spawning stock biomass at time *t* is then calculated as a function of
the numbers at age, the maturity at age, and the weight at age scaled by
a parameter $\gamma$. When $\gamma$ is greater than 1, the species is
said to experience hyperallometric fecundity, i.e. fecundity increases
faster than weight.

@@

Recruitment follows Beverton-Holt dynamics. We do however allow for five
variants in the timing of density dependence:

1.  Global density dependence: Density dependence is a function of the
    sum of spawning biomass across all patches, and recruits are then
    distributed according to habitat quality

$$
n_{t,p,s,a = 1} = \left(\frac{0.8{\times}\sum_{p=1}^P{r0_{p}}\times{h}\times\sum_{p=1}^P{ssb_{t-1,p}}}{0.2\times{\sum_{p=1}^Pssb0_p}\times(1 - h)+(h - 0.2)\times{\sum_{p=1}^Pssb_{t-1,p}}}\right)\times {hab_p}
$$ {#eq-dd2}

where *hab* is a vector of relative recruitment habitat quality in each
patch.

$$
hab_p = r0_p / \sum_p^P{r0_p}
$$ {#eq-hab}

2.  Local density dependence: Density dependence occurs independently in
    each patch and recruits are retained in their home patch.

$$
n_{t,p,s,a = 1,p} = \left(\frac{0.8{\times}r0_{s,p}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p}}}\right)
$$ {#eq-dd1}

3.  Local density dependence then disperse: Density dependence occurs
    independently in each patch and recruits are then dispersed.

$$
  n_{t,p,s,a = 1,p} = \left(\frac{0.8{\times}r0_{s,p}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p}}}\right)\times \boldsymbol{d^l} \times \epsilon_t
$$ {#eq-dd3}

@@

4.  Post-dispersal density dependence: Larvae are distributed throughout
    the system, and then density dependence occurs based on the density
    of adult biomass at the destination patch.

$$
  larv_{t,p} = ssb_{t-1} \times\boldsymbol{d^l}
$$ {#eq-larvmove}

$$
n_{t,a = 1,p} = \left(\frac{0.8{\times}r0_{p}\times{h}\times{larv_{t,p}}}{0.2\times{ssb0_p}\times(1 - h)+(h - 0.2)\times{larv_{t,p}}}\right) \times \epsilon_t
$$ {#eq-dd4}

@@

$$
n_{t,p,s,a = 1} = \left(\frac{0.8{\times}\sum_{p=1}^P{r0_{p}}\times{h}\times\sum_{p=1}^P{ssb_{t-1,p}}}{0.2\times{\sum_{p=1}^Pssb0_p}\times(1 - h)+(h - 0.2)\times{\sum_{p=1}^Pssb_{t-1,p}}}\right)\times \frac{ssb_{t-1,p}}{\sum_{p=1}^Pssb_{t-1,p}}
$$ {#eq-dd5}

@@

1.  *Rate*: MPAs are placed based on the pre-MPA
    depletion-weighted catch relative to the total catch in a patch. So,
    patches with high rates of catch of depleted species relative to
    total catch are prioritized.

@@

To our knowledge, MPAs are never randomly placed; their locations are
decided based on stakeholder negotiation given available information and
objectives. In theory, the design of MPA networks can be optimized
through the use of a modeling framework, and depending on the
validity of the model, this process is likely to produce better
outcomes than manually-designed strategies [@rassweiler2012;
@rassweiler2014]. However, designing optimal MPA networks becomes
increasingly difficult as the range of objectives and the complexity of
the model increase. Therefore, while `marlin` provides some support for
[optimization](https://danovando.github.io/marlin/articles/optimize-mpas.html),
we focus here on the design and performance of these more "rule of
thumb" design strategies that may be more accessible to a wider range of
users. We allow all MPAs to be designed in a "mosaic" fashion, but users
can easily extend the analysis to compare outcomes between
"contiguous" (MPA is made up of one continuously connected block) and "mosaic" (MPAs can be separated in space) MPA designs [@pons2022].

@@

In our coral reef example, we model the dynamics of four species: a
grouper (*Epinephelus fuscoguttatus*), a shallow-reef snapper (*Lutjanus
malabaricus*), a deep-water snapper (*Pristipomoides filamentosus*), and
a reef shark (*Carcharhinus amblyrhynchos*). The simulated groupers
undergo a mass migration to a spawning aggregation once per year,
followed by the sharks. Shallow-reef snappers stay in reefs closer to
shore year-round, while deep-water snappers stay in the deeper reefs
(@fig-coral).

These species are targeted by two different fleets. Fleet one primarily
targets the grouper and nearshore snapper populations, but will land any
incidentally captured sharks. Fleet one has a dome shaped selectivity
pattern for both grouper and shallow-reef snapper due to demand for
"plate sized" fish [@kindsvater2017], and logistic selectivity for
sharks, as any shark caught is retained. This fleet is a subsistence
fleet, where the local community takes advantage of every possible opportunity to
fish, and as such we model it as a "constant effort"
fishery.

The second fleet primarily targets the deep-water snapper population.
While plate-sized snapper are still prized from this population, we
model fleet two's selectivity as logistic due to high levels of discard
mortality for deep-water snapper resulting from barotrauma. Fleet two
catches some grouper, though less than fleet one, and receives no price
for sharks due to the requirements of a certification program through
which they sell their deep-water snapper. Accidental captures (bycatch) do
occur, which results in mortality. This fleet operates under open-access
dynamics, as fishing is not the primary means of subsistence;
short-term effort expands and contracts in response to profitability of
the primarily grouper-driven fishery.

@@

We use `marlin` to simulate the outcomes for both food production and
conservation for each of the species and both of the fleets as a
function of both MPA size and MPA design strategy. For this exercise,
MPAs are placed with perfect information and have no design constraints
for continuity (i.e. the selected MPA design can be as patchwork as
dictated by the data). We ran the simulation on a quarterly level ($\Delta_t = 1/4$) and set the area of each cell to an arbitrary value of 1 ($\Delta_p = 1$).

@@

critter_biomass %>%
  ungroup() %>% 
  left_join(sci_to_com, by = "critter") %>%
  select(-critter) %>%
  rename(critter = common_name) %>%
  filter(step == max(step)) %>% 
  ggplot(aes(reorder(critter, depletion), depletion)) +
  geom_col() +
  scale_y_continuous(limits = c(0, 1), name = "Depletion", expand = expansion(mult = c(0.0,0.05))) +
  labs(x = "") + 
  coord_flip()

@@

All code and materials needed to reproduce this paper is available at (https://github.com/DanOvando/marlin-paper)[https://github.com/DanOvando/marlin-paper].

>>>>

Abstract: Designing effective spatial management strategies is challenging because marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond endogenously to ecosystem changes in ways that depend upon ecological and policy context. We present a tool -- marlin -- to efficiently simulate the bio-economic dynamics of marine systems, designed to support both management and research. We demonstrate `marlin’s` capabilities by focusing on two case studies on the conservation and food production impacts of marine protected areas (MPAs): a coastal coral reef and a pelagic tuna fishery. In the coastal coral reef example, we explore how heterogeneity in species distributions and fleet preferences can affect distributional outcomes of MPAs. In the pelagic case study, we demonstrate how our model can be used to assess the climate resilience of different MPA design strategies, as well as the climate sensitivity of different fishing fleets. This paper demonstrates how intermediate complexity simulation of coupled bio-economic dynamics can help communities predict and potentially manage trade-offs between conservation, fisheries yields, and distributional outcomes resulting from spatial management.

@@

We define a few commonly used terms here. *Yields* refers to the volume of catch from fishing activity. *Spawning stock biomass* refers to the total biomass of reproductively mature fish in the water, a function of numbers, weight, and sexual maturity at age. We measure the size of the population by th ratio of the *spawning stock biomass* (SSB) in a given time step relative to *unfished spawning stock biomass* (SSB0), SSB/SSB0. So, a SSB/SSB0 value of 1 means that the population is unfished, a SSB/SSB0 value of 0 means the population is extinct.

@@

Revenues for each fleet in a time step are a function of the total amount of each species caught and its price ($\Pi$). The amount caught is a function of the selectivity at age of each species for each fleet ($\alpha$), the fishing efficiency of the fleet for that that species (also called "catchability", *q*), the amount of fishing effort of the fleet in question in that patch in a given time step (*E*), and the total instantaneous fishing mortality (*u*) at age (*a*) (including all other fleets) for that particular species in that patch and time step.

@@

Total catches or yield *Y* are calculated through the Baranov equation [@baranov1918], which appropriately accounts for the total instantaneous mortality (both fishing and natural. *z*) and divides the total amount of the population biomass (*b*) killed between natural (*m*) and fishing (*u*) sources, with the amount of biomass killed through fishing called "yield".

@@

This means that each species experiences a total mortality at age in a time step in a patch, individual fractions of which are portioned off as catches and subsequently revenues for each fleet. Given revenues *R*, we then calculate profits $\phi$ based on revenues and costs. Costs *C* are calculated as a function of base costs per unit effort ($\gamma$) as well as potential additional costs per unit effort of fishing in particular patches ($\eta$) for fleet *f*. $\beta$ allows for the cost per unit effort effort to scale non-linearly. Travel costs ($\eta$) are calculated based on the Euclidean distance of each patch to the nearest port of a given fleet, and users can specify any number from zero to the number of patches of port locations. When no ports are specified, travel costs are zero.

@@

#### Tuning Fishing Fleet Parameters

The degree of fishing pressure exerted by a given set of fishing fleets on each species is a function of a range of parameters including the total amount of effort (*E*), the selectivity ogives ($\alpha$,the curve specifying the proportion of individuals at a given age or size that will be caught by a unit of fishing mortality), fishing cost (*C*), the spatial distribution of the species affected by the fleet, the relative prices across species ($\Pi$), and the catchability coefficient of each fleet for each species ($q$). `marlin` provides a tuning option to help users achieve desired biological outcomes from their fleets.

Users can tune their fleets in one of two ways. First, they can specify a target exploitation rate *u* for each species in their simulation. Taking all the other parameters of the model as given, `marlin` then adjusts the catchability coefficients $q_{f,s}$ for each fleet *f* and species *s* such that the equilibrium exploitation rate for each species matches the desired level. Second, they can specify a target total spawning stock biomass under fishing divided by total unfished spawning stock biomass, and the model will adjust the catchability coefficients $q_{f,s}$ for each fleet *f* and species *s* such that the equilibrium ratio of fished to unfished spawning biomass for each species matches the desired level.

Users can also tune the fleet dynamics by specifying the ratio of costs to revenues. Price data for use in the model can be obtained relatively simply through literature reviews, market surveys, or local experts, so long as they are converted to the same units of biomass used in the model. However, cost parameters are more complicated, as translating say cost per day of fishing into the same representation of fishing effort used in the model is not straight forward to accomplish. As an alternative, users can specify an equilibrium cost-to-revenue ratio for the fleet, essentially the profit margins of the fleet in question, and `marlin` will tune the cost parameter to achieve this desired cost to revenue ratio given the other parameters in the model (e.g. price and biomass of the target species). So for example, while users may not have access to exact cost parameters, they may have a general sense that when operating under ideal conditions a given fleet has a profit margin of 10%, i.e. the cost to revenue ratio is 90%. They can input the cost to revenue ratio as 0.9 then and allow `marlin` to tune the cost parameters accordingly.

@@

The underlying population model used is an age structured single-species model in the manner of @ovando2021a. The population model requires many parameters. However, if the user supplies either a scientific or common name for the species in question (scientific preferred), the model will supply default values for that species based on the values reported in FishLife [internet connection required, @thorson2020]. FishLife provides estimates of core life history parameters for fish species based on a model integrating published values and evolutionary connections. FishLife provides reasonable default values for given species, but these default values should be not be taken as definitive and users should input best available parameters specific to the stock in question if they wish to best represent the dynamics of their specific system, and should check the plausibility of the default values prior to use. All of the life history values can be replaced with user-supplied values as needed and should be for more tactical applications; the default values simply ensure that reasonable and internally consistent values for the species in question are supplied as a starting point.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain (CTMC), as described in @thorson2021a. Within this framework, the model allows for movement to be broken down into three components of *advection* (follow currents), and *taxis* (active movement towards preferred habitat) and *diffusion* (essentially remaining variation in movement not explained by advection or taxis). For now, `marlin` focuses just on the diffusion and taxis components of this model, assuming that advection is zero, though future extensions could easily incorporate advection vectors from oceanographic models. In this way, `marlin` allows users to run anything from a simple Gaussian dispersal kernel up to a system governed by species that passively diffuse out from a core habitat defined by a dynamic thermal range.

We provide a brief overview of the the general CTMC method here, see @thorson2021a for a detailed description. Under this framework, movement from each patch to each other patch in the system in a given timestep *t* for life stage *a* of species *s* is defined by a movement matrix $\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the matrix exponent of the sum of the diffusion $\boldsymbol{D}$ and taxis $\boldsymbol{\tau}$ matrices scaled by the width of the time step $\Delta_{t}$ and the length of the edge of each patch $\Delta_d$ that the model is running under. So for example, if we are running a model on a monthly timestep, and we define the units of time in years, then $\Delta_{t} = 1/12$. Similarly, if we are running a model where one patch equals 100km^2^, then $\Delta_d = 100km$. This parameterization allows users to set the effective area of the spatial domain through two avenues; the number of patches, which effectively scales the resolution of the model, and the area of each patch, which scales the spatial extent of the simulation.

@@

The individual components of the diffusion matrix are filled based on an adjacency matrix, which defines whether two patches are both adjacent and water (as opposed to land or another physical barrier), and a diffusion rate $D$ defined in units of length of the side of a patch per unit of time. For example, if we are defining the time units as years and the distance units as kilometers, for a tuna $D$ might be 1,000 $\frac{KM}{Year}$. The $\Delta_{t}$ and $\Delta_{d}$ parameters then translate the movement rate $D$ to match the time step and patch size used in a simulation. This "scale free" parameterization means that appropriate value of $D$ can be identified for a species and then set, regardless of the time step or patch size used in the simulation model itself. These parameters are then used to calculate the value of a particular entry the diffusion matrix (*d*).

@@

The taxis matrix $\boldsymbol{\tau}$ is built off of the difference in log habitat quality between each patch (the habitat gradient), the adjacency matrix, and temporal and spatial resolution of the model ($\Delta_{t}$ and $\Delta_{p}$), as well as a scalar term $\lambda$. $\lambda$ scales the magnitude of the habitat gradient between patches relative to the diffusion parameter $D$. When the diffusion and taxis matrices are estimated jointly from actual data in the manner of @thorson2021a, $\lambda$ should be set to zero. However, when users are manually creating both the diffusion and taxis matrices for simulation, the relative strength of diffusion to taxis must be defined by the user. Setting higher values of $\lambda$ allows users to simulate a scenario where taxis is much more significant than diffusion.

$$
\tau_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}}(1 + \lambda d_{s,f})e^{log(h(p2,t,s,a) - log(h(p1,t,s,a))} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} \tau_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

For both the diffusion and taxis matrix, we allow for the inclusion of physical barriers to movement (i.e. land). Pairs of patches that are adjacent but in which one or both patches are a barrier to movement are set as non-adjacent. The CTMC model then produces movement dynamics that move around barriers rather than over them. This is different than setting habitat in a patch to zero; an animal will not preferentially move towards a patch with zero habitat all else being equal, but can move over a patch with zero habitat towards another better patch. An example might be a species that lives and feeds on coastal habitats transiting through a patch of open ocean to reach a new feeding ground. In contrast, a physical barrier represents something like a peninsula that an animal must move around to transit from one side to another.

While this CTMC approach to movement simulation is to date not commonly seen in the marine modeling literature, it has numerous advantages that warrant its seeming complexity. First, when only diffusion is present, the model will generalize to the familiar dynamics of a Gaussian dispersal kernel. However, the time component *t* in @eq-movement allows the movement model to be run on arbitrarily fine time steps. Second, the taxis model allow for more active habitat choices by species, allowing us to simulate both preferences of species in space and time efficiently. Lastly, this form has the advantage that the parameters are directly estimable from real data. So, if provided with say spatial abundance data and a tagging study, users can estimate the diffusion and taxis movement parameters in the same manner as @thorson2021a, and then pass their estimated parameters to `marlin` for simulation.

@@

where *BH* is the Beverton-Holt recruitment function [@beverton1957] and *ssb* is spawning-stock-biomass. Per convention, the model allows for a "plus group", wherein rather than tracking numbers of every possible age, individuals greater than or equal to a given maximum age are grouped together.

Spawning stock biomass *ssb* is calculated by converting age to mean length, calculating weight at age, maturity at age, and then calculating spawning stock biomass as the sum of spawning potential at age in a given time step, taking into account the potential for hyperallometry. Age is converted to length through the von Bertalanffy growth equation given parameters asymptotic length ($l_\infty$), growth (*k*) and theoretical age at length zero ($a0$) assuming log-normally distributed variation *u* in the length at age with CV $\sigma_s$.

@@

@@

Biomass *b* at age is then given by the weight at length equation governed by a scaling coefficient $\Omega_s$ and an exponent $\Phi_s$ that controls the rate at which volume scales with length

$$
  b_{a,s} = _s \times l_{a,s}^{wb_s}
$$ {#eq-weight}

The proportion of sexually mature individuals (*mat*) at a given age is then calculated as a logistic function where $l_{mat}$ is the length at which on average 50% of individuals are sexually mature, and $\delta_{mat}$ is the unit of length beyond $l_{mat}$ at which on average 95% of fish are sexually mature.

@@

Spawning stock biomass at time *t* is then calculated as a function of the numbers at age, the maturity at age, and the weight at age raised by a parameter $\gamma$. When $\gamma$ is greater than 1, the species is said to experience hyperallometric fecundity, i.e. fecundity increases faster than weight.

@@

Recruitment (i.e. the number of age 1 individuals entering the population) follows Beverton-Holt dynamics parameterized around steepness (*h*) with log-normally distributed recruitment deviates $\epsilon$. When steepness is one recruitment is independent of spawning biomass. As steepness approaches 0.2 recruitment becomes a linear function of spawning biomass. We allow for five variants in the timing of density dependent recruitment:

1.  Global density dependence: Density dependent recruitment is a function of the sum of spawning biomass across all patches, and recruits are then distributed according to habitat quality

$$
n_{t,p,s,a = 1} = \left(\frac{0.8{\times}\sum_{p=1}^P{r0_{p,s}}\times{h_s}\times\sum_{p=1}^P{ssb_{t-1,p,s}}}{0.2\times{\sum_{p=1}^Pssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{\sum_{p=1}^Pssb_{t-1,p,s}}}\right)\times {r0_{p,s} / \sum_p^P{r0_{p,s}}\times \epsilon_{t,s}
}
$$ {#eq-dd2}

where $r0$ is is a vector of recruits under unfished conditions in a given patch.

2.  Local density dependence: Density dependent recruitment occurs independently in each patch and recruits are retained in their home patch.

$$
n_{t,p,s,a = 1} = \left(\frac{0.8{\times}r0_{p,s}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p,s}}}\right) \times \epsilon_{t,s}
$$ {#eq-dd1}

3.  Local density dependence then disperse: Density dependent recruitment occurs independently in each patch and recruits are then dispersed.

$$
  n_{t,p,s,a = 1} = \left(\frac{0.8{\times}r0_{p,s}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p,s}}}\right)\times \boldsymbol{d^l_s} \times \epsilon_{t,s}
$$ {#eq-dd3}

@@

4.  Post-dispersal density dependence: Larvae are distributed throughout the system, and then density dependent recruitment occurs based on the density of adult biomass at the destination patch.

$$
  larv_{t,p,s} = ssb_{t-1,p,s} \times\boldsymbol{d^l_s}
$$ {#eq-larvmove}

$$
n_{t,a = 1,p,s} = \left(\frac{0.8{\times}r0_{p,s}\times{h_s}\times{larv_{t,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{larv_{t,p,s}}}\right) \times \epsilon_{t,s}
$$ {#eq-dd4}

@@

$$
n_{t,p,s,a = 1} = \left(\frac{0.8{\times}\sum_{p=1}^P{r0_{p,s}}\times{h_s}\times\sum_{p=1}^P{ssb_{t-1,p,s}}}{0.2\times{\sum_{p=1}^Pssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{\sum_{p=1}^Pssb_{t-1,p,s}}}\right)\times \frac{ssb_{t-1,p,s}}{\sum_{p=1}^Pssb_{t-1,p,s}} \times \epsilon_{t,s}
$$ {#eq-dd5}

Log-normal recruitment deviates are calculated with the potential for autocorrelation defined with strength $\rho$

$$ 
log(\epsilon_{t,s}) \sim \begin{cases}
       N(0,\sigma_{r,s}), & \text{if $t = 1$}\\
       \rho_s log(\epsilon_{t-1,s}) + \sqrt{1 - \rho_s^2} N(0,\sigma_{r,s}), & \text{if $t > 1$}
  \end{cases}
$$ {#eq-recdev}

And log recruitment deviates are converted to raw units using the bias correction factor

$$
\epsilon_{t,s} = e^{log(\epsilon_{t,s}) - \sigma_{r,s}^2/2}
$$ {#eq-bias}

@@

1.  *Rate*: MPAs are placed based on the pre-MPA SSB/SSB0 weighted catch relative to the total catch in a patch. So, patches with high rates of catch of depleted species relative to total catch are prioritized.

@@

To our knowledge, MPAs are never randomly placed; their locations are decided based on stakeholder negotiation given available information and objectives. In theory, the design of MPA networks can be optimized through the use of a modeling framework, and depending on the validity of the model, this process is likely to produce better outcomes than manually-designed strategies [@rassweiler2012; @rassweiler2014]. However, designing optimal MPA networks becomes increasingly difficult as the range of objectives and the complexity of the model increase. Therefore, while `marlin` provides some support for [optimization](https://danovando.github.io/marlin/articles/optimize-mpas.html), we focus here on the design and performance of these more "rule of thumb" design strategies that may be more accessible to a wider range of users. We allow all MPAs to be designed in a "mosaic" fashion, but users can easily extend the analysis to compare outcomes between "contiguous" (MPA is made up of one continuously connected block) and "mosaic" (MPAs can be separated in space) MPA designs [@pons2022] by developing and passing their own MPA design algorithms to the model.

@@

In our coral reef example, we model the dynamics of four tropical Pacific species: a grouper (*Epinephelus fuscoguttatus*), a shallow-reef snapper (*Lutjanus malabaricus*), a deep-water snapper (*Pristipomoides filamentosus*), and a reef shark (*Carcharhinus amblyrhynchos*). The simulated groupers undergo a mass migration to a spawning aggregation once per year, followed by the sharks. Shallow-reef snappers stay in reefs closer to shore year-round, while deep-water snappers stay in the deeper reefs (@fig-coral).

These species are targeted by two different fleets. Fleet One primarily targets the grouper and nearshore snapper populations, but will land any incidentally captured sharks. Fleet One has a dome shaped selectivity pattern for both grouper and shallow-reef snapper due to demand for "plate sized" fish [@kindsvater2017], and logistic selectivity for sharks, as any shark caught is retained. Fleet One is totally dependent on fishing for their livelihood, meaning the local community takes advantage of every possible opportunity to fish, and as such we model it as a "constant effort" fishery.

Fleet Two primarily targets the deep-water snapper population. While plate-sized snapper are still prized from this population, we model Fleet Two's selectivity as logistic due to high levels of discard mortality for deep-water snapper resulting from barotrauma. Fleet Two catches some grouper, though less than Fleet One, and receives no price for sharks due to the requirements of a certification program through which they sell their deep-water snapper. Accidental captures (bycatch) do occur, which results in mortality. Fleet Two operates under open-access dynamics, as fishing is not the primary means of subsistence; short-term effort expands and contracts in response to profitability of the primarily grouper-driven fishery.

@@

We use `marlin` to simulate the outcomes for both food production and conservation for each of the species and both of the fleets as a function of both MPA size and MPA design strategy. For this exercise, MPAs are placed with perfect information and have no design constraints for continuity (i.e. the selected MPA design can be as patchwork as dictated by the data). We ran the simulation on a quarterly level ($\Delta_t = 1/4$) and set the area of each patch to an arbitrary value of 1 ($\Delta_p = 1$).

@@

critter_biomass %>%
  ungroup() %>% 
  left_join(sci_to_com, by = "critter") %>%
  select(-critter) %>%
  rename(critter = common_name) %>%
  filter(step == max(step)) %>% 
  ggplot(aes(reorder(critter, depletion), depletion)) +
  geom_col() +
  scale_y_continuous(limits = c(0, 1), name = "SSB/SSB0", expand = expansion(mult = c(0.0,0.05))) +
  labs(x = "") + 
  coord_flip()

@@

All code and materials needed to reproduce this paper is available at (https://github.com/DanOvando/marlin-paper)\[https://github.com/DanOvando/marlin-paper\].
